---
title: "ddct"
author: "Kat Valero"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
#comments

##Lamprey function for use on any plate
#Reads and cleans the plate sample sheet from a QuantStudio thermocycler, removes blanks, parses gene and sample IDs, and assigns technical replicates.
#Reads raw LAMP fluorescence data, converts signal columns to numeric, and selects the chosen fluorescence channel.
#Calculates a forward 5-cycle signal ratio per well to identify the cycle of strongest amplification.
#Applies a minimum amplification threshold and returns inferred iCT values joined to the sample metadata.

#libraries
```{r}
library(dplyr)
library(stringr)
library(purrr)
library(tibble)
```

#iCt5
```{r}


compute_plate_iCT_n5 <- function(samples_file, raw_file, plate_no, thr = 1.2, channel = "x1-m1") {

  # sample sheet
  samples <- read.csv(samples_file, skip = 46, header = TRUE, check.names = FALSE) %>%
    rename(well = `Well`, well_pos = `Well Position`, sample_name = `Sample Name`) %>%
    mutate(
      well_pos = trimws(as.character(well_pos)),
      sample_name = trimws(as.character(sample_name)),
      sample_name = na_if(sample_name, ""),              # <-- remove empty strings
      plate_no = plate_no
    ) %>%
    filter(!is.na(sample_name), tolower(sample_name) != "blank") %>%  # <-- removes blanks too
    mutate(
      gene = str_trim(str_remove(sample_name, "\\s*#.*$")),
      lamp_sample = as.integer(str_extract(sample_name, "\\d+$"))
    ) %>%
    group_by(gene, lamp_sample) %>%
    arrange(well_pos, .by_group = TRUE) %>%
    mutate(techrep = row_number()) %>%
    ungroup()

  # raw sheet
  raw <- read.csv(raw_file, skip = 46, header = TRUE, check.names = FALSE) %>%
    rename(well = `Well`, well_pos = `Well Position`) %>%
    mutate(
      well_pos = trimws(as.character(well_pos)),
      plate_no = plate_no
    )

  # numeric convert x*-m* columns
  xcols <- grep("^x\\d+-m\\d+$", names(raw), value = TRUE)
  raw[xcols] <- lapply(raw[xcols], function(z) as.numeric(gsub(",", "", z)))

  raw$signal <- raw[[channel]]

  eps <- 1e-6

  raw_lags <- raw %>%
    group_by(well_pos) %>%
    arrange(Cycle, .by_group = TRUE) %>%
    mutate(norm_n5 = lead(signal, 5) / pmax(signal, eps)) %>%
    ungroup()

  ict_by_well <- raw_lags %>%
    group_by(well_pos) %>%
    summarise(
      iCT_n5 = Cycle[which.max(ifelse(is.na(norm_n5), -Inf, norm_n5))],
      max_n5 = max(norm_n5, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      iCT_n5 = ifelse(is.infinite(max_n5), NA, iCT_n5),
      max_n5 = ifelse(is.infinite(max_n5), NA, max_n5),
      iCT_n5 = ifelse(max_n5 < thr, NA, iCT_n5)
    )

  out <- samples %>% left_join(ict_by_well, by = "well_pos")
  out
}


```

#iCt1
```{r}
compute_plate_iCT_n1 <- function(samples_file, raw_file, plate_no, thr = 1.2, channel = "x1-m1") {

   samples <- read.csv(samples_file, skip = 46, header = TRUE, check.names = FALSE) %>%
    rename(well = `Well`, well_pos = `Well Position`, sample_name = `Sample Name`) %>%
    mutate(
      well_pos = trimws(as.character(well_pos)),
      sample_name = trimws(as.character(sample_name)),
      sample_name = na_if(sample_name, ""),
      plate_no = plate_no
    ) %>%
    filter(!is.na(sample_name), tolower(sample_name) != "blank") %>%
    mutate(
      gene = str_trim(str_remove(sample_name, "\\s*#.*$")),
      lamp_sample = as.integer(str_extract(sample_name, "\\d+$"))
    ) %>%
    group_by(gene, lamp_sample) %>%
    arrange(well_pos, .by_group = TRUE) %>%
    mutate(techrep = row_number()) %>%
    ungroup()

  raw <- read.csv(raw_file, skip = 46, header = TRUE, check.names = FALSE) %>%
    rename(well = `Well`, well_pos = `Well Position`) %>%
    mutate(
      well_pos = trimws(as.character(well_pos)),
      plate_no = plate_no
    )

  xcols <- grep("^x\\d+-m\\d+$", names(raw), value = TRUE)
  raw[xcols] <- lapply(raw[xcols], function(z) as.numeric(gsub(",", "", z)))

  raw$signal <- raw[[channel]]
  eps <- 1e-6

  raw_lags <- raw %>%
    group_by(well_pos) %>%
    arrange(Cycle, .by_group = TRUE) %>%
    mutate(norm_n1 = lead(signal, 1) / pmax(signal, eps)) %>%
    ungroup()

  ict_by_well <- raw_lags %>%
    group_by(well_pos) %>%
    summarise(
      iCT_n1 = Cycle[which.max(ifelse(is.na(norm_n1), -Inf, norm_n1))],
      max_n1 = max(norm_n1, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      iCT_n1 = ifelse(is.infinite(max_n1), NA, iCT_n1),
      max_n1 = ifelse(is.infinite(max_n1), NA, max_n1),
      iCT_n1 = ifelse(max_n1 < thr, NA, iCT_n1)
    )

  out <- samples %>% left_join(ict_by_well, by = "well_pos")
  out
}

```

#ict 2
```{r}
compute_plate_iCT_n2 <- function(samples_file, raw_file, plate_no, thr = 1.2, channel = "x1-m1") {

  library(dplyr)
  library(stringr)

  samples <- read.csv(samples_file, skip = 46, header = TRUE, check.names = FALSE) %>%
    rename(well = `Well`, well_pos = `Well Position`, sample_name = `Sample Name`) %>%
    mutate(
      well_pos = trimws(as.character(well_pos)),
      sample_name = trimws(as.character(sample_name)),
      sample_name = na_if(sample_name, ""),
      plate_no = plate_no
    ) %>%
    filter(!is.na(sample_name), tolower(sample_name) != "blank") %>%
    mutate(
      gene = str_trim(str_remove(sample_name, "\\s*#.*$")),
      lamp_sample = as.integer(str_extract(sample_name, "\\d+$"))
    ) %>%
    group_by(gene, lamp_sample) %>%
    arrange(well_pos, .by_group = TRUE) %>%
    mutate(techrep = row_number()) %>%
    ungroup()

  raw <- read.csv(raw_file, skip = 46, header = TRUE, check.names = FALSE) %>%
    rename(well = `Well`, well_pos = `Well Position`) %>%
    mutate(
      well_pos = trimws(as.character(well_pos)),
      plate_no = plate_no
    )

  xcols <- grep("^x\\d+-m\\d+$", names(raw), value = TRUE)
  raw[xcols] <- lapply(raw[xcols], function(z) as.numeric(gsub(",", "", z)))

  raw$signal <- raw[[channel]]
  eps <- 1e-6

  raw_lags <- raw %>%
    group_by(well_pos) %>%
    arrange(Cycle, .by_group = TRUE) %>%
    mutate(norm_n2 = lead(signal, 2) / pmax(signal, eps)) %>%
    ungroup()

  ict_by_well <- raw_lags %>%
    group_by(well_pos) %>%
    summarise(
      iCT_n2 = Cycle[which.max(ifelse(is.na(norm_n2), -Inf, norm_n2))],
      max_n2 = max(norm_n2, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      iCT_n2 = ifelse(is.infinite(max_n2), NA, iCT_n2),
      max_n2 = ifelse(is.infinite(max_n2), NA, max_n2),
      iCT_n2 = ifelse(max_n2 < thr, NA, iCT_n2)
    )

  out <- samples %>% left_join(ict_by_well, by = "well_pos")
  out
}

```


##run the function across all plates and merge the results 
#Identifies all sample-sheet and raw-data CSV files in a folder and extracts plate numbers from filenames.
#Matches sample and raw files by plate number to build a per-plate lookup table.
#Iterates over plates and applies compute_plate_iCT() to each, combining results into a single data frame.
#Writes the combined per-sample iCT results to disk for downstream analysis.

```{r}


folder <- "C:/directory/lamp_ict"

samples_files <- list.files(folder, pattern = "_samples\\.csv$", full.names = TRUE)
raw_files     <- list.files(folder, pattern = "_raw\\.csv$",     full.names = TRUE)

get_plate_no <- function(path) {
  as.integer(str_extract(tolower(basename(path)), "\\d+(?=_(samples|raw)\\.csv$)"))
}

samples_tbl <- tibble(
  samples_file = samples_files,
  plate_no = get_plate_no(samples_files)
) %>%
  filter(!is.na(plate_no)) %>%
  distinct(plate_no, .keep_all = TRUE)

raw_tbl <- tibble(
  raw_file = raw_files,
  plate_no = get_plate_no(raw_files)
) %>%
  filter(!is.na(plate_no)) %>%
  distinct(plate_no, .keep_all = TRUE)

plates <- inner_join(samples_tbl, raw_tbl, by = "plate_no") %>%
  arrange(plate_no)

# --------- RUN n = 1 ----------
all_iCT_n1 <- pmap_dfr(
  list(plates$samples_file, plates$raw_file, plates$plate_no),
  ~ compute_plate_iCT_n1(..1, ..2, ..3, thr = 1.2, channel = "x1-m1")
)

# --------- RUN n = 2 ----------
all_iCT_n2 <- pmap_dfr(
  list(plates$samples_file, plates$raw_file, plates$plate_no),
  ~ compute_plate_iCT_n2(..1, ..2, ..3, thr = 1.2, channel = "x1-m1")
)

# --------- RUN n = 5 ----------
all_iCT_n5 <- pmap_dfr(
  list(plates$samples_file, plates$raw_file, plates$plate_no),
  ~ compute_plate_iCT_n5(..1, ..2, ..3, thr = 1.2, channel = "x1-m1")
)

# --------- RENAME COLUMNS SO THEY DON'T CLASH ----------
all_iCT_n1 <- all_iCT_n1 %>% rename(iCT_n1 = iCT_n1)
all_iCT_n2 <- all_iCT_n2 %>% rename(iCT_n2 = iCT_n2)
all_iCT_n5 <- all_iCT_n5 %>% rename(iCT_n5 = iCT_n5)

# --------- MERGE THEM ----------
all_samples_with_all_iCTs <- all_iCT_n1 %>%
  select(plate_no, gene, lamp_sample, techrep, iCT_n1) %>%
  left_join(
    all_iCT_n2 %>% select(plate_no, gene, lamp_sample, techrep, iCT_n2),
    by = c("plate_no", "gene", "lamp_sample", "techrep")
  ) %>%
  left_join(
    all_iCT_n5 %>% select(plate_no, well_pos, gene, lamp_sample, techrep, iCT_n5),
    by = c("plate_no", "gene", "lamp_sample", "techrep")
  )

# --------- SAVE ----------
write.csv(
  all_samples_with_all_iCTs,
  file.path(folder, "all_plates_iCT_n1_n2_n5.csv"),
  row.names = FALSE
)

```

#Compute the FDM with base R (stats package) and add to the table with iCT values

```{r}

folder <- "C:/directory/lamp_ict"

# --- files already used for iCT ---
samples_files <- list.files(folder, pattern = "_samples\\.csv$", full.names = TRUE)
raw_files     <- list.files(folder, pattern = "_raw\\.csv$",     full.names = TRUE)

get_plate_no <- function(path) {
  as.integer(str_extract(tolower(basename(path)), "\\d+(?=_(samples|raw)\\.csv$)"))
}

samples_tbl <- tibble(
  samples_file = samples_files,
  plate_no = get_plate_no(samples_files)
) %>%
  filter(!is.na(plate_no)) %>%
  distinct(plate_no, .keep_all = TRUE)

raw_tbl <- tibble(
  raw_file = raw_files,
  plate_no = get_plate_no(raw_files)
) %>%
  filter(!is.na(plate_no)) %>%
  distinct(plate_no, .keep_all = TRUE)

plates <- inner_join(samples_tbl, raw_tbl, by = "plate_no") %>%
  arrange(plate_no)

#  FDM fit for one well (4PL sigmoid)
fit_fdm_one <- function(dat) {

  dat <- dat[order(dat$Cycle), ]
  x <- dat$Cycle
  y_raw <- dat$signal

  # baseline correction (first 10 cycles)
  base <- mean(y_raw[x <= 10], na.rm = TRUE)
  y <- y_raw - base
  y[y < 0] <- 0

  # skip if essentially flat
  if (max(y, na.rm = TRUE) < 1000) {
    return(data.frame(FDM = NA, fit_ok = FALSE))
  }

  A0 <- min(y, na.rm = TRUE)
  K0 <- max(y, na.rm = TRUE)
  M0 <- x[which.min(abs(y - (A0 + (K0 - A0)/2)))]
  B0 <- 0.2

  tryCatch({
    fit <- nls(
      y ~ A + (K - A) / (1 + exp(-B * (x - M))),
      start = list(A = A0, K = K0, B = B0, M = M0),
      control = nls.control(maxiter = 200)
    )
    co <- coef(fit)
    data.frame(FDM = unname(co["M"]), fit_ok = TRUE)
  }, error = function(e) {
    data.frame(FDM = NA, fit_ok = FALSE)
  })
}

# compute FDM per plate from *_raw.csv
compute_plate_FDM <- function(raw_file, plate_no, channel = "x1-m1", max_cycle = 80) {

  raw <- read.csv(raw_file, skip = 46, header = TRUE, check.names = FALSE) %>%
    rename(well_pos = `Well Position`) %>%
    mutate(
      well_pos = trimws(as.character(well_pos)),
      plate_no = plate_no
    )

  # numeric convert x*-m* columns
  xcols <- grep("^x\\d+-m\\d+$", names(raw), value = TRUE)
  raw[xcols] <- lapply(raw[xcols], function(z) as.numeric(gsub(",", "", z)))

  raw <- raw %>%
    mutate(signal = raw[[channel]],
           Cycle = as.numeric(Cycle)) %>%
    filter(Cycle <= max_cycle)

  fdm_tbl <- raw %>%
    group_by(plate_no, well_pos) %>%
    group_modify(~ fit_fdm_one(.x)) %>%
    ungroup()

  fdm_tbl
}

all_FDM <- pmap_dfr(
  list(plates$raw_file, plates$plate_no),
  ~ compute_plate_FDM(..1, ..2, channel = "x1-m1", max_cycle = 80)
)

write.csv(all_FDM,
          file.path(folder, "all_plates_FDM_from_raw.csv"),
          row.names = FALSE)

# merge into the iCT table 
ict_all <- read.csv(file.path(folder, "all_plates_iCT_n1_n2_n5.csv"))

ict_plus_fdm <- ict_all %>%
  left_join(all_FDM, by = c("plate_no", "well_pos"))

write.csv(ict_plus_fdm,
          file.path(folder, "all_plates_iCT_n1_n2_n5_FDM.csv"),
          row.names = FALSE)

```


#Loads master LAMP dataset and inferred iCT values, standardises keys, and merges iCT and FDM into the main data table containing auto-called Ct values.
```{r}


#  Load master data (contains CT) 
df <- read.csv("C:directory/sv_lampdata.csv", check.names = FALSE)

if ("Plate_#" %in% names(df)) df <- df %>% rename(plate = `Plate_#`)

df_clean <- df %>%
  mutate(
    plate = as.integer(str_extract(as.character(plate), "\\d+")),
    `LAMP_sample_#` = as.integer(str_extract(as.character(`LAMP_sample_#`), "\\d+")),
    techrep = as.integer(str_extract(as.character(techrep), "\\d+")),
    gene = as.character(gene),
    CT = as.numeric(gsub(",", "", as.character(CT)))
  )

# Load new iCT + FDM table
ict_fdm <- read.csv("C:/directory/all_plates_iCT_n1_n2_n5_FDM.csv")

ict_fdm_clean <- ict_fdm %>%
  mutate(
    plate = as.integer(str_extract(as.character(plate_no), "\\d+")),
    `LAMP_sample_#` = as.integer(str_extract(as.character(lamp_sample), "\\d+")),
    techrep = as.integer(str_extract(as.character(techrep), "\\d+")),
    gene = as.character(gene)
  )

# Merge everything
merged_all <- df_clean %>%
  left_join(
    ict_fdm_clean %>%
      select(plate, `LAMP_sample_#`, techrep, gene,
             iCT_n1, iCT_n2, iCT_n5, FDM, fit_ok),
    by = c("plate", "LAMP_sample_#", "techrep", "gene")
  )

write.csv(merged_all,
          "C:/directory/SV_CT_iCT_FDM_per_well.csv",
          row.names = FALSE)


```

